Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Addressed comments of reviewer C.

Going beyond the standard syntax-directed equality check, there is a technique that can deal with dynamic $\eta$-expansion.  It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the computed normal forms for identity and, thus, decide definitional equality.
NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

@@

\item We present the first integration of a dependent type theory with higher-rank size polymorphism.
Concretely, we consider a type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).  Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
(In the following rules, $a$, $a'$, and $b$ stand for arbitrary size expressions.)
\[
\ru{\Gamma \der t = t' : \forall i.\, T
  }{\Gamma \der t\,a = t\,a' : T[b/i]}
\qquad
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[b/i]}
\qquad
\qquad
\ru{\Gamma \der t = t' : \Pi i.\, T
  }{\Gamma \der t\,a = t\,a : T[a/i]}
\qquad
\ru{\Gamma \der t : \Pi i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
Our substitution theorem distinguishes term- from type-side substitutions.

@@

As already done in some examples, we may use a named dependent
function type notation as syntactic sugar for the corresponding de
Bruijn representation.  For instance,
$\funT z \Size \Nat z \to \Set[\ell]$ is sugar for
$\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this
type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
$\forall z.\ (\sfunT x {\Nat z} T\, z\, x) \to \sfunT x {\Nat (z + 1)} T \,(z+1)\, x$.
%
Similar as for $\Pi$, we use named lambda abstraction as sugar for de
Bruijn abstraction.  Named abstraction takes care of proper lifting of
de Bruijn indices, for instance,
$\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the
scope of $x$.
%
We may also use names when we construct concrete contexts, for
instance, if\/ $T$ is well-formed in context $\Gamma$, we may write
$T\,z\,x$ in context $\Gamma.z\of\Size.x\of\Nat z$ to mean
$T\slift^2\,\ind 1\,\ind 0$ in context $\Gamma.\Size.\Nat \ind 0$.

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
We have \fbox{boxed} the rules dealing with irrelevant size application.
Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
annotations $T$ in recursion $\fix[\ell] T t$.  The type constructor $T$
has to be monotone in the size argument; this is a technical condition
for type-based termination \cite{gimenez:typeBased}.  We will make use
of it in Section~\ref{sec:comp}.

@@

\emph{Size values} \fbox{$\alpha, \beta \in \SIZE$} are size
expressions that use de Bruijn levels instead of de Bruijn
indices. Comparison of size values $\alpha \leq \beta$ is
analoguous to comparison of size terms $a \leq b$.  In the following, we
will reuse letter $a$ for a value if it cannot be confused for a size
term.

@@

\begin{lemma}[Subsumption]
\label{lem:subsump} \hfill
\begin{enumerate}
\item \label{it:natsub} If\/ $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
\item \label{it:allsub} If\/ $\F(\alpha) \subseteq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
\item \label{it:pisub}  If\/ $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
\item \label{it:usub}  If\/ $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
\end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof}
  Propositions (\ref{it:natsub}--\ref{it:pisub}) are clear.
  For (\ref{it:usub}), prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
  For the base types this is direct, let us look at the function space.
\[
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \qquad
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
\]
  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
\end{proof}
}
We define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
Simultaneously, we need to prove correctness, namely that $T \leq T' \in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.  The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
\begin{gather*}
  \ru{T = T' \in \NEE
    }{T \leq T' \in \TYPE}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      T' \evalsto \Nat \alpha' \qquad
      \alpha \leq \alpha'
    }{T \leq T' \in \TYPE}
\qquad
  \ru{T \evalsto \Set[\ell_0] \qquad
      T' \evalsto \Set[\ell_0'] \qquad
      \ell_0 \leq \ell_0'
    }{T \leq T' \in \TYPE}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      T' \evalsto \piT {A'}{B'} \qquad
      A' \leq A \in \TYPE \qquad
      B[u] \leq B'[u'] \in \TYPE \mforall u = u' \in A'
    }{T \leq T' \in \TYPE}
\\[2ex]
  \ru{T \evalsto \epiT \Size B \qquad
      T' \evalsto \epiT {\Size}{B'} \qquad
      B[\alpha] \leq B'[\alpha] \in \TYPE \mforall \alpha \in \SIZE
    }{T \leq T' \in \TYPE}
\end{gather*}
\begin{lemma}[Subtyping is preorder]
\label{lem:preord}
\bla
\begin{enumerate}
\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T' \in \TYPE$.
\item If\/ $T_1 \leq T_2 \in \TYPE$ and $T_2 \leq T_3 \in \TYPE$ then $T_1 \leq T_3 \in \TYPE$.
\end{enumerate}
\end{lemma}

@@

Reflection and reification perform $\eta$-expansion so that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.

@@

\begin{theorem}[Reflection and reification]
\label{thm:rere}
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If\/ $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If\/ $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof*}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

\nextcase $T \evalsto \piT A B$ with $A \in \SET[\ell]$ and
      $B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A$
\begin{gather*}
%   \ru{A \in \SET[\ell] \qquad
%       B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
%     }{T \in \SET[\ell]}
% \\[2ex]
  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
      A \shape A_1 \qquad
      B[u] \shape B_1[u'] \mforall u = u' \in A
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
      A \shape A_2 \qquad
      B[u] \shape B_2[u'] \mforall u = u' \in A
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
  Let $d_i = \down{A_i}{u_i}$.
  By induction hypothesis, $d_1 = d_2 \in \NF$.
  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$ by Lemma~\ref{lem:closne}, and
  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.

\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
It is sufficient to show
$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
\end{enumerate}

@@

yet, are not general enough for function types.

@@

By definitions, these relations are closed under subsumption and

@@

In this article, we have described the first successful integration of higher-rank size polymorphism into a core type theory with dependent function types, a sized type of natural numbers, a predicative hierarchy of universes, subtyping, and $\eta$-equality.  This is an important stepping stone for the smooth integration of sized types into dependently-typed proof assistants.
In the final paragraphs, we discuss some questions and insights that follow from our work and go beyond it.

@@

% Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types, and even principal types.

>>>>

Going beyond the standard syntax-directed equality check, there is a
technique that can deal with dynamic $\eta$-expansion.  It is a
type-directed normalization function inspired by
normalization-by-evaluation (NbE) that computes $\eta$-long normal
forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the
computed normal forms for identity and, thus, decide definitional
equality.  NbE has proven to be a robust method to decide equality in
powerful type theories with non-trivial $\eta$-laws.  It scales to
universes and large eliminations \cite{abelCoquandDybjer:lics07},
topped with singleton types or proof irrelevance
\cite{abelCoquandPagano:lmcs11}, and even impredicativity
\cite{abel:flops10}.  At its heart there are reflection $\up T$ and
reification $\down T$ functions directed by type $T$ and orchestrating
just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables
$x$ into the realm of values of type $T$ and lets us compute with
open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and
computes its long normal form.  For instance, the normal form of a
closed function $f : U \to T$ would be
$\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed
variant $f : \funT x U T[x]$ it would be
$\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

@@

\item We present the first integration of a dependent type theory with higher-rank size polymorphism.
Concretely, we consider a type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).  Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
(In the following rules, $a$, $a'$, and $b$ stand for arbitrary size expressions.)
\[
\ru{\Gamma \der t = t' : \forall i.\, T
  }{\Gamma \der t\,a = t'\,a' : T[b/i]}
\qquad
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[b/i]}
\qquad
\qquad
\ru{\Gamma \der t = t' : \Pi i.\, T
  }{\Gamma \der t\,a = t'\,a : T[a/i]}
\qquad
\ru{\Gamma \der t : \Pi i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
Our substitution theorem distinguishes term- from type-side substitutions.

@@

As already done in some examples, we may use a named dependent
function type notation as syntactic sugar for the corresponding de
Bruijn representation.  For instance,
$\funT z \Size \Nat z \to \Set[\ell]$ is sugar for
$\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this
type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for
$\forall z.\ (\sfunT x {\Nat z} T\, z\, x) \to \sfunT x {\Nat (z + 1)} T \,(z+1)\, x$.
%
Similarly to for $\Pi$, we use named lambda abstraction as sugar for de
Bruijn abstraction.  Named abstraction takes care of proper lifting of
de Bruijn indices, for instance,
$\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the
scope of $x$.
%
We may also use names when we construct concrete contexts, for
instance, if\/ $T$ is well-formed in context $\Gamma$, we may write
$T\,z\,x$ in context $\Gamma.z\of\Size.x\of\Nat z$ to mean
$T\slift^2\,\ind 1\,\ind 0$ in context $\Gamma.\Size.\Nat \ind 0$.

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
We have \fbox{boxed} the rules dealing with irrelevant size application.
Judgement $\Gamma \der T : \Adm \ell$ characterizes the valid type
annotations $T$ in recursion $\fix[\ell] T t$.  The type constructor $T$
has to be monotone in the size argument; this is a technical condition
for type-based termination \cite{gimenez:typeBased}.  We will make use
of it in Section~\ref{sec:comp}.
We write \fbox{$\DD :: J$} to express that $\DD$ is a derivation of judgement $J$.

@@

\emph{Size values} \fbox{$\alpha, \beta \in \SIZE$} are size
expressions that use de Bruijn levels instead of de Bruijn
indices. Comparison of size values $\alpha \leq \beta$ is
analogous to comparison of size terms $a \leq b$.  In the following, we
will reuse letter $a$ for a value if it cannot be confused for a size
term.

@@

\begin{lemma}[Subsumption]
\label{lem:subsump} \hfill
\begin{enumerate}
\item \label{it:natsub} If\/ $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
\item \label{it:allsub} If\/ $\F(\alpha) \subseteq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
\item \label{it:pisub}  If\/ $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
\item \label{it:usub}  If\/ $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
\end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof}
  Propositions (\ref{it:natsub}--\ref{it:pisub}) are clear.
  For (\ref{it:usub}), prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
  For the base types this is direct, let us look at the function space.
\[
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \qquad
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
\]
  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
\end{proof}
}
We define subtyping of type values \fbox{$T \leq T' \in \TYPE$} by induction on $T \in \SET[\ell]$ and $T' \in \SET[\ell']$.
Simultaneously, we need to prove correctness, namely that $T \leq T' \in \TYPE$ implies $\EL(T) \subseteq \EL(T')$.  The correctness follows from Lemma~\ref{lem:subsump} and we do not spell it out here.
\begin{gather*}
  \ru{T = T' \in \NEE
    }{T \leq T' \in \TYPE}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      T' \evalsto \Nat \alpha' \qquad
      \alpha \leq \alpha'
    }{T \leq T' \in \TYPE}
\qquad
  \ru{T \evalsto \Set[\ell_0] \qquad
      T' \evalsto \Set[\ell_0'] \qquad
      \ell_0 \leq \ell_0'
    }{T \leq T' \in \TYPE}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      T' \evalsto \piT {A'}{B'} \qquad
      A' \leq A \in \TYPE \qquad
      B[u] \leq B'[u'] \in \TYPE \mforall u = u' \in A'
    }{T \leq T' \in \TYPE}
\\[2ex]
  \ru{T \evalsto \epiT \Size B \qquad
      T' \evalsto \epiT {\Size}{B'} \qquad
      B[\alpha] \leq B'[\alpha] \in \TYPE \mforall \alpha \in \SIZE
    }{T \leq T' \in \TYPE}
\end{gather*}
\begin{lemma}[Subtyping is a preorder]
\label{lem:preord}
\bla
\begin{enumerate}
\item If\/ $T = T' \in \SET[\ell]$ then $T \leq T' \in \TYPE$.
\item If\/ $T_1 \leq T_2 \in \TYPE$ and $T_2 \leq T_3 \in \TYPE$ then $T_1 \leq T_3 \in \TYPE$.
\end{enumerate}
\end{lemma}

@@

Reflection and reification perform $\eta$-expansion so that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, the precise type is not needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.

@@

\begin{theorem}[Reflection and reification]
\label{thm:rere}
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item \label{it:up}
If\/ $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item \label{it:down}
If\/ $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof*}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

\nextcase $T \evalsto \piT A B$ with $A \in \SET[\ell]$ and
      $B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A$
\begin{gather*}
%   \ru{A \in \SET[\ell] \qquad
%       B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
%     }{T \in \SET[\ell]}
% \\[2ex]
  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
      A \shape A_1 \qquad
      B[u] \shape B_1[u'] \mforall u = u' \in A
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
      A \shape A_2 \qquad
      B[u] \shape B_2[u'] \mforall u = u' \in A
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
  Let $d_i = \down{A_i}{u_i}$.
  By induction hypothesis (\ref{it:down}) with shapes $A \shape A_1$ and $A \shape A_2$
  we get $d_1 = d_2 \in \NF$.
  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$ by Lemma~\ref{lem:closne}, and
  by induction hypothesis (\ref{it:up}) with shapes
  $B[u_1] \shape B_1[u_1]$ and $B[u_1] \shape B_2[u_2]$ we obtain
  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.

\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis, since $\var k = \var k \in \NE$ by Lemma~\ref{lem:closne}.
It is sufficient to show
$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
\end{enumerate}

@@

but need to be strengthened for function types.

@@

By definition, these relations are closed under subsumption and

@@

In this article, we have described the first successful integration of higher-rank size polymorphism into a core type theory with dependent function types, a sized type of natural numbers, a predicative hierarchy of universes, subtyping, and $\eta$-equality.  This is an important stepping stone for the smooth integration of sized types into dependently-typed proof assistants.
In these final paragraphs, we discuss some questions and insights that follow from our work and go beyond it.

@@

% Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.

In the long run, we wish for a type-directed equality check that does
not normalization in one go, but interleaves weak head normalization
with structural comparison.
Such a equality test is at the heart of Agda's type
checker and it generates constraints for meta variables
involved in type reconstruction \cite{norell:PhD}.  However, the usual
bidirectional construction \cite{abelScherer:types10} does not seem to
go through as we lack uniqueness of types (and even principal types).
